### 2021-12-14 input/proposal.tex
reworked detection and followup text

@@

\begin{definition}[\label{def:fs}\underline{F}orwarding of
  \underline{s}taleness (FS)]
  %
  Any source can broadcast a staleness notification at any
  time. Every \process $x$ delivers and forwards a received staleness
  notification if it targets its best up-to-date delivered message:
  %% \TODO{$(\delta_s \in R_x \wedge \alpha_s =
  %%     \textup{minUTD}~D_x) \iff \eventually (d_x(\delta_s) \rightarrow
  %% f_x(\delta_s))$.}
  $(\textup{last}~D_x = \alpha_s \wedge d_x(\alpha_s) \rightarrow
  r_x(\delta_s)) \implies d_x(\delta_s) \rightarrow f_x(\delta_s)$.
  % \wedge \nexists d'_s: d_x(\delta_s) \TODO{\not\rightarrow} d_x(\delta_s'))$.
\end{definition}

\begin{definition}[\label{def:fbplus}$\textup{FB}^+$: echos] FB where
  a \process $x$ that receives but does not deliver a staleness
  notification sends back --~or echoes~-- its best up-to-date
  delivered message: $(\textup{last}~D_x = \alpha_s^d \wedge \delta_s
  \not\in D_x \wedge d_x(\alpha_s^d)\rightarrow r_{xy}(\delta_{s'}))
  \implies r_{xy}(\delta_{s'}) \rightarrow
  s_{xy}(\alpha_s^d\oplus^{w_{xy}})$.

@@

To reach consistent partitioning, \processes need
\begin{inparaenum}[(i)]
\item to detect possibly blocking conditions
\item to purge all stale messages from the system
\item enabling up-to-date notifications to eventually echo back to
  every \process.
\end{inparaenum}
Since Figure~\ref{fig:problemD} highlights that when the protocol
terminates, $c$ is the \process that remains in the wrong partition,
we put the burden of detection and resolution on it.

\TODO{A lot of blockings occur, only that of example is a problem.
  Could be $b$ or $c$ that detects? We put burden on $c$ since
  unconsistency starts here. Problem: how to make sure that $c$
  receives a staleness notification in this case? Also, $b$ is not
  sure to receive $\delta_c$ since composable problem.}

\begin{lemma}[\label{lem:detector}$\textup{FB}^+ \wedge \textup{FS}
  \implies$ \underline{D}etector existence (D)] Assuming
  $\textup{FB}^+$, $\textup{FS}$, a chain of delivery on $\pi_{xz}$ of
  $\alpha_x$ with $\alpha_x = \textup{last}~D_z$, if $x$ broadcasts a staleness
  notification $\delta_x$, either $z$ eventually delivers it, or a
  \process $y$ in such a chain of delivery $\pi_{xz}'$ eventually
  acknowledges its blocking.
  % $b_p(\delta_x) \implies d_q(x) U Detect[X]$ ??? 
\end{lemma}

\begin{proof}
  %% knowing that b_p delta is sufficient to say that processes will
  %% receive delta or detect blocking
  (A) Assuming a chain of delivery $\pi_{xz} = [x, \ldots, z]$ of
  $\alpha_x$ with $\forall y \in \pi_{xz}: \alpha_x = \min D_y$, and
  assuming $d_x(\delta_x)$, then with forwarding of staleness
  (Definition~\ref{def:fs}),
  $f_{\pi_{xz}[k]}(\delta_x) \implies \eventually
  d_{\pi_{xz}[k+1]}(\delta_x)$ except if $\delta_x \in D_{\pi_{xz}[k+1]}$ which
  means that $\pi_{xz}[k+1]$ already delivered and forwarded $\delta_x$
  following another such a chain of delivery. Therefore, $z$
  eventually receives, hence delivers $\delta_x$.

  \noindent (B) Assuming a chain of delivery
  $\pi_{xz} = [x, \ldots, z]$ of $\alpha_x$ with at least
  $\alpha_x = \min D_z$, and assuming $d_x(\delta_x)$, then by (A),
  the staleness notification reaches a \process $ y = \pi_{xz}[k]$. By
  Definition~\ref{def:fs}, the forwarding stops when $y$ already
  delivered $\delta_x$ ($\delta_x \in D_y$ covered by (A)) or
  delivered a better message $\alpha_s = \min D_y$ with
  $\alpha_s \neq \alpha_x$. With $y'=\pi_{xz}[k+1]$, forwarding of
  best (Definition~\ref{def:fbplus}) implies three possible outcomes:
  \begin{enumerate}[(i)]
  \item $\alpha_{s'} = \min D_{y'}$ with $\alpha_{s'} \neq \alpha_x$
    hence $y' \iff y$ which leaves two possible outcomes:
  \item
    $r_{y'}(\alpha_s) \wedge \alpha_x = \min D_{y'} \wedge \alpha_x <
    \alpha_s$ which means that a shorter path of delivery $\pi_{xz}'$
    exists that either forwards appropriate staleness notifications
    (A) or detects its possible blocking:
  \item
    $r_{y'}(\alpha_s) \wedge \delta_s \in D_{y'} \wedge \alpha_x =
    \min D_{y'}$ which detects a possible blocking of
    $\delta_x$. Without global knowledge, $y'$ assumes it belongs to
    the shortest and only path of delivery of $\alpha_x$
    ($\pi_{xy'} \subseteq \Pi_{xz}$) thus it does not further delegate
    the detection to another \process.
  \end{enumerate} \vspace{-1.5em} %% ugly vspace but necessary for enumerate+qed
  %
  %% (i) If every \process (such as $a$) starting from the source
  %% delivers and forwards a removal notification $\delta_X$ when their
  %% last delivery is $\alpha_X$, then every such \process eventually
  %% delivers the removal notification $\delta_X$ except \processes (such
  %% as $b$) that delivered $\alpha_X$ from a \process that delivered a
  %% message from another partition $\alpha_Y$ since then.
  %
  %% These exceptions eventually receive and deliver $\alpha_Y$ since
  %% $\alpha_Y$ is better than $\alpha_X$ through this path, except if
  %% they already received and delivered the removal notification of
  %% $P_Y$ (such as $c$). These \processes may never receive hence
  %% deliver $\delta_X$, and may never receive hence deliver a better
  %% message than $\alpha_X$. They need an additional mechanism to
  %% eventually purge $\alpha_X$ that cannot rely on the eventual purging
  %% of $P_Y$ at \processes like $b$, to avoid deadlocks.
  %
  %% (ii) Assuming that every \process keeps an history of its past
  %% deliveries, \processes (such as $c$) can detect the inconsistency,
  %% since they receive from their parent an already deleted partition
  %% $P_Y$. This notification means that the parent discards any
  %% $\delta_Z$ with $\alpha_Z^z < \alpha_Y^y$, and most importantly, if
  %% $\delta_X$ exists, it discards it. To ensure the purge of stale
  %% messages, a detecting \process must assume the worst case that such
  %% $\delta_X$ exists, and send another kind of message, noted $\Delta$,
  %% that notify the possible removal of $P_X$.
  %
  %% (iii) A \process (such as $d$) whose last delivery is $\alpha_X$,
  %% but whose parent is neither inconsistent (like $b$) nor receiving
  %% $\delta_X$ (like $a$), eventually receives $\Delta$ from a detecting
  %% \process, either directly or transitively, for such a child \process
  %% ($d$) trusts the possible removal of $P_X$ by delivering and
  %% forwarding such $\Delta$. $\Delta$ suffers from identical blocking
  %% conditions (between $b$ and $c$) than $\delta$, leading to the same
  %% solutions of detection and propagation of $\Delta$. Eventually,
  %% every \process whose last delivery is $\alpha_X$ receives and
  %% delivers either a better $\alpha_Z$, or $\delta_X$, or $\Delta_X$.
\end{proof}

Proof of Lemma~\ref{lem:detector} shows that there exists false
positive in the blocking detection: $\delta_x$ may not exist but $y'$
still receives $\alpha_s$ after the delivery of $\delta_s$ from the
path it delivered $\alpha_x$. Figure~\ref{fig:undoproblem} highlights
this behavior. In Figure~\ref{fig:undoproblemB}, $a$ does not delete
its partition. In Figure~\ref{fig:undoproblemD} and because of \NAMEB,
$c$ received the same series of messages as in
Figure~\ref{fig:problemD}. Yet, $c$ must assume the existence of
$\delta_a$ and act accordingly by forcing the staleness of its best
delivered message and disseminate this information to its neighbors.
To avoid flooding the system with false positive staleness
notifications, we propose to reduce the scope of staleness
notifications by including only downstream \processes. In
Figure~\ref{fig:undoproblemD}, false positives would not affect $a$
and $b$.

\begin{definition}[\label{def:fsplus}$\textup{FS}^+$: forwarding
    staleness downstream] Any \emph{\process} can broadcast a
  staleness notification at any time. A \emph{child \process} $x$
  delivers and forwards a received staleness notification if it comes
  from the path of delivery of its best up-to-date delivered
  message. \TODO{more.}
\end{definition}

@@

%% \NAME guarantees dynamic consistent partitioning by making extensive
%% use of \process-to-\process communications. It requires
%% \begin{inparaenum}[(i)]
%% \item purging stale control information, hence propagating $\delta$
%%   messages; and
%% \item retrieving the closest source, hence propagating $\alpha$
%%   messages.
%% \end{inparaenum}
\TODO{TODOUBLECHECK}.
% \TODO{Multiple receipt and
% multiple
%  delivery imply more messages (receipt bounded by $|O|$ as well).}
Deleting the $a^{th}$ partition generates twice as many messages as
creating the $a^{th}$ partition: $\delta$ messages travel through the
partition, then $\alpha$ messages compete to fill the gap.  Overall,
the communication complexity shows that \NAME scales well to the
number of partitions.
%% In the
%% worst-case, every new partition includes all but one \process
%% belonging to the previous partition. The total number of messages
%% after the $a^{th}$ new partition is $\mathcal{O}(\overline{|O|}\cdot
%% a^2)$. As for the average-case, the number of messages for the
%% partition deletion is identical to the number of messages of the
%% corresponding partition creation.
%% \begin{algorithm}[h!]
%%   \input{input/algoedges.tex}
%%   \caption{\label{algo:edges}\NAME at \Process~$p$ in dynamic networks.}
%% \end{algorithm}

>>>>

\newcommand{\last}{\textup{last}~}

@@

\begin{definition}[\label{def:fs}\underline{F}orwarding of
  \underline{s}taleness (FS)]
  %
  Any source can broadcast a staleness notification at any
  time. Every \process $x$ delivers and forwards a received staleness
  notification if it targets its best up-to-date delivered message:
  %% \TODO{$(\delta_s \in R_x \wedge \alpha_s =
  %%     \textup{minUTD}~D_x) \iff \eventually (d_x(\delta_s) \rightarrow
  %% f_x(\delta_s))$.}
  $(\last D_x = \alpha_s \wedge d_x(\alpha_s) \rightarrow
  r_x(\delta_s)) \implies d_x(\delta_s) \rightarrow f_x(\delta_s)$.
  % \wedge \nexists d'_s: d_x(\delta_s) \TODO{\not\rightarrow} d_x(\delta_s'))$.
\end{definition}

\begin{definition}[\label{def:fbplus}$\textup{FB}^+$: echos] FB where
  a \process $x$ that receives but does not deliver a staleness
  notification sends back --~or echoes~-- its best up-to-date
  delivered message: $(\last D_x = \alpha_s^d \wedge \delta_s
  \not\in D_x \wedge d_x(\alpha_s^d)\rightarrow r_{xy}(\delta_{s'}))
  \implies r_{xy}(\delta_{s'}) \rightarrow
  s_{xy}(\alpha_s^d\oplus^{w_{xy}})$.

@@

A first step to avoid staying in inconsistent state is to extend the
behavior of each \process so \processes such as $c$, that would remain
in a wrong partition, use their history of receipt and delivery to
detect the blocking of staleness notifications that can hinder
reaching consistent partitioning. For instance, $c$ detects the
blocking of $\delta_a$ because it acknowledges that its parent $b$
delivered a stale notification $\alpha_c^1$, \ie between the delivery
of $\alpha_c^1$ and the delivery of $\delta_c$, $b$ blocked any other
$\delta$.

%% \Begincomme{inparaenum}[(i)] \item to detect possibly blocking
%% conditions that can hinder \item to purge all stale messages from
%% the system \item enabling up-to-date notifications to eventually
%% echo back to every \process.  \end{inparaenum} Since
%% Figure~\ref{fig:problemD} highlights that when the protocol
%% terminates, $c$ is the \process that remains in the wrong
%% partition, we put the burden of detection and resolution on it.

%% \TODO{A lot of blockings occur, only that of example is a problem.
%%   Could be $b$ or $c$ that detects? We put burden on $c$ since
%%   unconsistency starts here. Problem: how to make sure that $c$
%%   receives a staleness notification in this case? Also, $b$ is not
%%   sure to receive $\delta_c$ since composable problem.}

\begin{lemma}[\label{lem:detector}$\textup{FB}^+ \wedge \textup{FS}
  \implies$ Detector existence] Assuming
  $\textup{FB}^+$ and $\textup{FS}$, a chain of delivery on $\pi_{xz}$ of
  $\alpha_x$ with $\alpha_x = \last D_z$, if $x$ broadcasts a staleness
  notification $\delta_x$, either $z$ eventually delivers it, or a
  \process $y$ in such a chain of delivery $\pi_{xz}'$ eventually
  acknowledges its blocking.
  % $b_p(\delta_x) \implies d_q(x) U Detect[X]$ ??? 
\end{lemma}

\begin{proof}
  %% knowing that b_p delta is sufficient to say that processes will
  %% receive delta or detect blocking
  (A) Assuming a chain of delivery $\pi_{xz} = [x, \ldots, z]$ of
  $\alpha_x$ with $\forall y \in \pi_{xz}: \alpha_x = \last D_y$, and
  assuming $d_x(\delta_x)$, then with forwarding of staleness
  (Definition~\ref{def:fs}),
  $f_{\pi_{xz}[k]}(\delta_x) \implies \eventually
  d_{\pi_{xz}[k+1]}(\delta_x)$ except if $\delta_x \in D_{\pi_{xz}[k+1]}$ which
  means that $\pi_{xz}[k+1]$ already delivered and forwarded $\delta_x$
  following another such a chain of delivery. Therefore, $z$
  eventually receives, hence delivers $\delta_x$.

  \noindent (B) Assuming a chain of delivery
  $\pi_{xz} = [x, \ldots, z]$ of $\alpha_x$ with at least
  $\alpha_x = \last D_z$, and assuming $d_x(\delta_x)$, then by (A),
  the staleness notification reaches a \process $ y = \pi_{xz}[k]$. By
  Definition~\ref{def:fs}, the forwarding stops when $y$ already
  delivered $\delta_x$ ($\delta_x \in D_y$ covered by (A)) or
  delivered a better message $\alpha_s = \last D_y$ with
  $\alpha_s \neq \alpha_x$. With $y'=\pi_{xz}[k+1]$, forwarding of
  best (Definition~\ref{def:fbplus}) implies three possible outcomes:
  \begin{enumerate}[(i)]
  \item $\alpha_{s'} = \last D_{y'}$ with $\alpha_{s'} \neq \alpha_x$
    hence $y' \iff y$ which leaves two possible outcomes:
  \item
    $r_{y'}(\alpha_s) \wedge \alpha_x = \last D_{y'} \wedge \alpha_x <
    \alpha_s$ which means that a shorter path of delivery $\pi_{xz}'$
    exists that either forwards appropriate staleness notifications
    (A) or detects its possible blocking:
  \item
    $r_{y'}(\alpha_s) \wedge \delta_s \in D_{y'} \wedge \alpha_x =
    \last D_{y'}$ which detects a possible blocking of
    $\delta_x$. Without global knowledge, $y'$ assumes it belongs to
    the shortest and only path of delivery of $\alpha_x$
    ($\pi_{xy'} \subseteq \Pi_{xz}$) thus it does not further delegate
    the detection to another \process.
  \end{enumerate} \vspace{-1.5em} %% ugly vspace but necessary for enumerate+qed
  %
  %% (i) If every \process (such as $a$) starting from the source
  %% delivers and forwards a removal notification $\delta_X$ when their
  %% last delivery is $\alpha_X$, then every such \process eventually
  %% delivers the removal notification $\delta_X$ except \processes (such
  %% as $b$) that delivered $\alpha_X$ from a \process that delivered a
  %% message from another partition $\alpha_Y$ since then.
  %
  %% These exceptions eventually receive and deliver $\alpha_Y$ since
  %% $\alpha_Y$ is better than $\alpha_X$ through this path, except if
  %% they already received and delivered the removal notification of
  %% $P_Y$ (such as $c$). These \processes may never receive hence
  %% deliver $\delta_X$, and may never receive hence deliver a better
  %% message than $\alpha_X$. They need an additional mechanism to
  %% eventually purge $\alpha_X$ that cannot rely on the eventual purging
  %% of $P_Y$ at \processes like $b$, to avoid deadlocks.
  %
  %% (ii) Assuming that every \process keeps an history of its past
  %% deliveries, \processes (such as $c$) can detect the inconsistency,
  %% since they receive from their parent an already deleted partition
  %% $P_Y$. This notification means that the parent discards any
  %% $\delta_Z$ with $\alpha_Z^z < \alpha_Y^y$, and most importantly, if
  %% $\delta_X$ exists, it discards it. To ensure the purge of stale
  %% messages, a detecting \process must assume the worst case that such
  %% $\delta_X$ exists, and send another kind of message, noted $\Delta$,
  %% that notify the possible removal of $P_X$.
  %
  %% (iii) A \process (such as $d$) whose last delivery is $\alpha_X$,
  %% but whose parent is neither inconsistent (like $b$) nor receiving
  %% $\delta_X$ (like $a$), eventually receives $\Delta$ from a detecting
  %% \process, either directly or transitively, for such a child \process
  %% ($d$) trusts the possible removal of $P_X$ by delivering and
  %% forwarding such $\Delta$. $\Delta$ suffers from identical blocking
  %% conditions (between $b$ and $c$) than $\delta$, leading to the same
  %% solutions of detection and propagation of $\Delta$. Eventually,
  %% every \process whose last delivery is $\alpha_X$ receives and
  %% delivers either a better $\alpha_Z$, or $\delta_X$, or $\Delta_X$.
\end{proof}

As a second step, detecting \processes must proactively purge the
system from their forwarded notifications. For instance in
Figure~\ref{fig:problemD}, $c$ could broadcast $\delta_a$ in order to
acknowledge the staleness of its own $\alpha_a^3$ and inform
neighboring \processes that delivered it as well. Unfortunately, proof
of Lemma~\ref{lem:detector} shows that there exists false positive in
the blocking detection: $\delta_x$ may not exist but $y'$ still
receives $\alpha_s$ after the delivery of $\delta_s$ from the path it
delivered $\alpha_x$. Figure~\ref{fig:undoproblem} highlights this
behavior. In Figure~\ref{fig:undoproblemB}, $a$ does not delete its
partition. In Figure~\ref{fig:undoproblemD} and because of scoped
broadcast, $c$ received the same series of messages as in
Figure~\ref{fig:problemD}. Yet, $c$ must assume the existence of
$\delta_a$ and act accordingly by forcing the staleness of its best
delivered message and disseminate this information to its neighbors.
To avoid flooding the system with false positive staleness
notifications, we propose to reduce the scope of staleness
notifications by including only downstream \processes. In
Figure~\ref{fig:undoproblemD}, false positives would not affect $a$
and $b$.

\begin{definition}[\label{def:fsplus}$\textup{FS}^+$: forwarding
    staleness downstream] Any \emph{\process} can broadcast a
  staleness notification at any time. A \emph{child \process} $x$
  delivers and forwards a received staleness notification if it comes
  from the path of delivery of its best up-to-date delivered
  message.
\end{definition}

@@

% \TODO{Multiple receipt and multiple delivery}
In terms of number of messages, deleting the $a^{th}$ partition
generates twice as many messages as creating the $a^{th}$ partition:
$\delta$ messages travel through the partition, then $\alpha$ messages
compete to fill the gap \TODO{which is?}. \TODO{What about false
  positives?} Overall, the communication complexity shows that \NAME
scales well to the number of partitions.
%% In the
%% worst-case, every new partition includes all but one \process
%% belonging to the previous partition. The total number of messages
%% after the $a^{th}$ new partition is $\mathcal{O}(\overline{|O|}\cdot
%% a^2)$. As for the average-case, the number of messages for the
%% partition deletion is identical to the number of messages of the
%% corresponding partition creation.
